From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-06-10 draft_v1/draft_v1.tex
update_draft_Leoniefeedback

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study},
  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{From Causal Discovery to Intervention Simulation: Modeling
Precarity and Depression in the HELIUS Study}

@@

\date{2025-05-20}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between domain-specific
precariousness indicators, and depressive symptoms. Our multi-resolution
analysis---spanning both aggregate and symptom-level
constructs---identified financial stress as a robust driver of
depression, and highlighted key symptoms, such as sleep disturbance and
depressed mood, as likely initiators or reinforcers of downstream
precariousness. To explore how these structural insights shape
intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains---including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025). This broader view reveals how different forms
of insecurity can co-occur and compound, forming an ecosystem of risk.
However, despite growing evidence of association between precariousness
and mental health, a fundamental question remains unresolved: \emph{How
do these factors causally influence one another?} Most existing studies
rely on cross-sectional correlations, leaving open questions about
directionality and feedback. Without stronger causal insight, it remains
difficult to identify priority targets for intervention or to anticipate
unintended effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precarity--depression relationship
affect the outcomes of stress-reducing interventions?} To answer this,
we combine two approaches. First, we use cycle-capable causal discovery
algorithms to infer directional relationships among domain-specific
precariousness factors, and depressive symptoms. By examining both
aggregated measures (depression sum score and a composite precarity
index) and disaggregated variables (individual depressive symptoms and
specific life-domain indicators), we capture broad patterns of
relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precarity co-evolve under varying levels of
external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. In some configurations, the system
exhibits bistability, remaining stuck in a high-risk state unless
external support exceeds a critical threshold. In others, it shows
unistable behavior, responding more gradually to incremental changes. By
integrating causal discovery with dynamic simulation, we aim to
demonstrate how the internal structure of the depression--precarity
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and which provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted exploratory analyses to identify
consistent and meaningful data-driven factor structures. These analyses
led to the identification of five distinct precariousness
factors---three domain-based and two reflecting recent stressors---each
composed of multiple observed variables, as outlined below. Additional
details on the HELIUS study and the factor extraction process are
provided in the \hyperref[sec-appendix]{Appendix}.

@@

After preprocessing, the HELIUS dataset includes 21,628 samples. In
addition to the five identified precariousness factors, we computed an
overall precarity score by aggregating their values. Depression is
represented using the PHQ-9 (Kroenke et al., 2001), both as a total sum
score and as individual symptom scores. In the subsequent causal
discovery analysis, we explore the relationship between depression and
precarity using both aggregated scores and their disaggregated
components. See Figure~\ref{fig-dist} for the overall distributions of
the variables used in the analysis.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay.\\
\emph{P.emp} = employment precariousness; \emph{P.hou} = housing
precariousness; \emph{P.soc} = social precariousness; \emph{S.fin} =
recent financial stressors; \emph{S.rel} = recent relational stressors;
\emph{Precarity} = overall precarity; \emph{PHQsum} = PHQ-9 sum score;
\emph{anh} = anhedonia; \emph{app} = appetite; \emph{con} =
concentration; \emph{dep} = depressed mood; \emph{ene} = energy;
\emph{glt} = guilty; \emph{mot} = motor; \emph{sui} = suicidal}

@@

\subsection{Modeling Dynamics between Precarity and
Depression}\label{modeling-dynamics-between-precarity-and-depression}

@@

NSGA-II is implemented via the \texttt{nsga2R} package (Franz \&
Nakamura, 2015), with a population of 500 evolved over 300 generations
using standard hyperparameters (crossover probability = 0.9, mutation
probability = 0.25, and tournament size = 2). This process yielded a
diverse ensemble of Pareto-efficient solutions balancing trade-offs
between objectives. From the resulting Pareto front
(Figure~\ref{fig-pareto}), we selected the top 30 parameter sets with
the lowest combined error to use in the subsequent simulations assessing
intervention effects.

@@

\subsection{Causal Structure Linking Precarity and
Depression}\label{causal-structure-linking-precarity-and-depression}

@@

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precariousness factors (\emph{P.hou},
\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
depression sum score (\emph{PHQsum}), as identified by the FCI
algorithm. As described in Section~\ref{sec-analysis}, dashed edges
indicate ties, with the exact edge endpoint where the tie occurs is
represented by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the
relationships between precarity factors and depression (\emph{PHQsum}).
Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
are not identified as causes of depression, whereas financial stress
(\emph{S.fin}) appears to play a causal role. While \emph{P.emp} and
\emph{P.soc} are generally not recognized as causes of other
precariousness factors, \emph{S.fin} emerges as a potential cause, as
indicated by its circle edge endpoint. Supporting this, the proportion
matrix plot in panel (b) shows that \emph{S.fin} has some probability of
causally influencing either \emph{P.emp} or \emph{P.soc.}

@@

Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
with depression and \emph{S.fin} through a latent confounder. Its
connection to \emph{P.soc} is less definitive---the proportion matrix in
Figure~\ref{fig-sum-2} indicates a considerable probability that this
relationship is either absent, ambiguous, or potentially bidirectional.
Meanwhile, housing precarity (\emph{P.hou}) is not directly causally
related to depression but is linked to \emph{P.emp} and \emph{S.fin}.
While \emph{P.emp} and \emph{S.fin} are identified as non-causes of
\emph{P.hou}, it remains unclear whether \emph{P.hou} causally
influences \emph{P.emp} or if their relationship is mediated by an
unobserved confounder.

@@

\caption{\label{fig-sum}Resulting graph of precarity factors and
depression sum score using FCI and proportion of edge endpoint types.}

@@

Despite these differences, some consistent patterns emerge across both
CI tests, particularly in the case of \emph{S.fin}, which appears to be
connected to nearly all other variables---though many of these
connections are marked by ties, reflecting uncertainty in
directionality. One exception is the stronger evidence suggesting that
\emph{S.fin} may cause changes in \emph{app}, while its relationships
with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
with depressive symptoms, with most edges pointing toward \emph{P.soc}
rather than outward from it. This suggests that depressive symptoms,
particularly \emph{dep} and \emph{glt}, may contribute to worsening
social precarity rather than the other way around. Additionally,
\emph{anh} also shows some probability of causally influencing
\emph{P.soc}, reinforcing the idea that social precarity is more often a
consequence rather than a driver of depressive symptoms.

@@

Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
to be the most connected to precariousness factors, while \emph{P.soc}
has the most connections with symptoms, predominantly as a recipient
rather than a driver of influence. Within the symptom network,
\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
symptoms, as they exhibit more outgoing arrows compared to other
symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
\emph{glt}, despite having high connectivity, predominantly receive
incoming arrows, indicating they are more likely effects rather than
causes. Considering both symptom-precariousness connections and
symptom-level dynamics, \emph{slp} emerges as a central symptom, given
its strong ties to precariousness factors and its influential role
within the symptom network. This suggests that sleep issues could be an
initiating symptom, particularly sensitive to relational stress and
employment precarity. On the other hand, \emph{P.soc} serves as a key
bridge between the depression and precariousness subsystems, as
depressive symptoms appear to feed back into social precarity,
reinforcing a self-sustaining dynamic between depression and precarious
conditions.

@@

\caption{\label{fig-sym}Resulting graph of precarity factors and
individual depression symptoms using FCI and proportion of edge endpoint
types.}

@@

\subsubsection{Precarity as sum score}\label{precarity-as-sum-score}

@@

Overall, the aggregated precariousness graph supports the key findings
from both the disaggregated analysis and the depression sum score
analysis, providing greater certainty in several important patterns. The
results highlight the significant role of latent confounders in
symptom-to-symptom interactions. While certain stressor-related
precariousness factors, particularly \emph{S.fin} and \emph{S.rel},
appear to causally influence depressive symptoms, other precariousness
factors are more likely driven by depression rather than acting as
primary causes, suggesting that precariousness may function more as a
consequence of depression than its source. Some symptoms, particularly
\emph{glt}, \emph{slp}, and \emph{dep}, show greater sensitivity to
precarious conditions, while also playing a central role in symptom
dynamics. Among them, \emph{slp} and \emph{glt} emerge as influential
symptoms within the depression network, whereas \emph{con} and
\emph{dep}, despite their high connectivity, behave more as outcomes
rather than primary drivers. Together, these findings suggest that
depression---especially symptoms like sleep disturbances and guilt---may
contribute to sustaining precarious conditions, possibly forming a
reciprocal cycle between depression and precariousness.

@@

\caption{\label{fig-presum}Resulting graph of precarity sum score and
individual depression symptoms using FCI and proportion of edge endpoint
types.}

@@

\subsection{Simulating the Dynamics Between Precarity and Depression
Under
Intervention}\label{simulating-the-dynamics-between-precarity-and-depression-under-intervention}

To evaluate how internal system structure influences the effect of
external support, we simulated outcomes under a range of intervention
strengths across the top 30 parameter sets from the calibrated model
ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
levels of depression (\(D\)) and precarity (\(P\)) after simulation,
with each line representing one parameter set. Lines are color-coded by
combined feedback and input strength, calculated as the product
\(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).

@@

Tailoring intervention strategies to the system's underlying structure
may enhance their effectiveness and sustainability, particularly in
addressing the compounding effects of socioeconomic precarity on mental
health.

@@

\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
precarity (\(P\)) across varying intervention strengths, simulated over
the top 30 parameter sets. Each line represents one parameter set and is
colored by its combined feedback and input strength.} \textbf{Bottom}:
Smoothed Density Heatmap of Final \(D\) and \(P\) for Two Parameter
Sets. \emph{One exhibits unistability (top two panels) and the other
exhibits bistability (bottom two panels). Bistable dynamics reveal a
sharp transition between distinct outcome regimes, while unistable
dynamics show gradual, continuous change across the intervention
gradient.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precarity
interact---and how their internal structure modulates responsiveness to
intervention. Using cross-sectional data from the HELIUS study, we
applied cycle-capable discovery algorithms to estimate plausible
directional relationships among multiple domains of precariousness
factors, and depressive symptoms. By analyzing both aggregate constructs
and individual variables, we aimed to capture patterns that may be
masked in either extreme of abstraction (i.e., purely aggregate or
disaggregated analyses).

The causal analysis consistently identified financial stress as a key
plausible driver of depression, appearing across both aggregated and
disaggregated models. At the symptom level, we found that sleep
disturbance, depressed mood, guilt, and anhedonia were especially
sensitive to external stressors. Among these, the symptom depressed mood
emerged as a central hub: not only was it heavily influenced by other
symptoms, it also exerted directional influence on social
precariousness, suggesting a feedback loop in which worsening mood
contributes to social disconnection, which in turn may exacerbate
depressive symptoms. The CCI algorithm even suggested a possible cyclic
relationship between depressed mood and overall precariousness, hinting
at a structurally self-reinforcing mechanism. Symptom sleep disturbance
also played a key initiating role. It showed directed connections to
both social and employment precariousness and exhibited many outgoing
edges within the symptom network. This suggests the symptom sleep
disturbance may function as an early warning signal or entry point into
the depression--precariousness cycle, making it a promising target for
preventive intervention. In contrast, Social precariousness more often
appeared as an effect rather than a cause---receiving arrows from
symptoms like depressed mood and guilt, but rarely sending them. This
asymmetry implies that social precariousness may often reflect
downstream consequences of worsening mental health rather than being its
root cause. These patterns offer actionable insights for intervention.
Addressing symptoms like sleep disturbance early may prevent escalation
into broader depressive states or worsening precarity. Intervening on
depressed mood, as a central node with potential feedback effects, could
also disrupt the self-reinforcing dynamics between mental health and
social isolation.

To examine how such structures shape responsiveness to support, we
developed a simplified two-variable nonlinear dynamical model capturing
the feedback between depression and precarity, modulated by financial
stress. Calibrated against empirical summary statistics from the HELIUS
dataset, the model allowed us to simulate how these dynamics evolve
under varying levels of stress reduction. The simulations revealed that
intervention effectiveness depends not just on the magnitude of external
stressors, but on the internal configuration of the system itself. In
bistable regimes---characterized by strong feedback and low noise---the
system could remain stuck in a high-risk state unless an intervention
crosses a critical threshold, tipping it into a healthier equilibrium.
By contrast, unistable systems---driven by weaker coupling or greater
noise---responded more gradually and proportionally to increasing
support. These dynamics carry practical policy implications. In bistable
systems, even moderate interventions may have little effect unless they
are forceful enough to tip the system toward a lower-risk state. This
underscores the importance of early or intensive support when targeting
self-reinforcing cycles of depression and precarity. Unistable systems,
by contrast, are more responsive to incremental support, suggesting that
sustained, moderate interventions may be sufficient in such contexts.

@@

Finally, the structure of the dynamical model introduces its own
limitations. While the model was designed to reflect key features of the
discovered causal structure, it represents a highly simplified system:
only two variables---depression and social precariousness---modulated by
external financial stress. This abstraction allowed for tractable
analysis and visual intuition, particularly around nonlinear behaviors
such as bistability, but omits other potentially influential constructs
highlighted in the causal graphs, including relational stress,
employment conditions and housing conditions. Moreover, the model's
nonlinearity---introduced via the hyperbolic tangent function---was
chosen for convenience rather than empirical justification. While this
function supports saturation and the emergence of tipping points, the
presence of bistability is a product of simulation, not empirical
verification. It remains possible that simpler, linear dynamics could
account for the observed data just as well. Future work should further
investigate whether bistability is a necessary feature of
depression--precarity dynamics, and explore alternative functional
forms---both linear and nonlinear---using richer data and stronger
empirical constraints.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study},
  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{From Causal Discovery to Intervention Simulation: Modeling
Precariousness and Depression in the HELIUS Study}

@@

\date{2025-06-10}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between domain-specific
precariousness indicators, and depressive symptoms. Our multi-resolution
analysis---spanning both aggregate and symptom-level
constructs---identified financial stress as a robust driver of
depression, and highlighted key symptoms, such as sleep disturbance and
depressed mood, as likely initiators or reinforcers of downstream
precariousness. To explore how these structural insights shape
intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precariousness under varying levels of financial stress reduction.
Simulation results revealed that the system's responsiveness depends not
only on the intensity of external intervention but also on internal
structural features such as feedback strength and noise. In particular,
some configurations exhibited bistability, requiring interventions to
surpass a critical threshold to trigger lasting improvement, while
others responded more proportionally. Together, these findings emphasize
that internal feedback dynamics---not just external adversity---can
shape the impact of mental health interventions. More broadly, our study
offers a scalable framework for integrating causal discovery and
dynamical simulation to better understand and inform policy responses to
complex psychosocial phenomena.
\end{abstract}

@@

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains---including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
how different forms of insecurity can co-occur and compound, forming an
ecosystem of risk. However, despite growing evidence of association
between precariousness and mental health, a fundamental question remains
unresolved: \emph{How do these factors causally influence one another?}
Most existing studies rely on cross-sectional correlations, leaving open
questions about directionality and feedback. Without stronger causal
insight, it remains difficult to identify priority targets for
intervention or to anticipate unintended effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precariousness--depression
relationship affect the outcomes of stress-reducing interventions?} To
answer this, we combine two approaches. First, we use cycle-capable
causal discovery algorithms to infer directional relationships among
domain-specific precariousness factors, and depressive symptoms. By
examining both aggregated measures (depression sum score and a composite
precariousness index) and disaggregated variables (individual depressive
symptoms and specific life-domain indicators), we capture broad patterns
of relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precariousness co-evolve under varying
levels of external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. In some configurations, the system
exhibits bistability, remaining stuck in a high-risk state unless
external support exceeds a critical threshold. In others, it shows
unistable behavior, responding more gradually to incremental changes. By
integrating causal discovery with dynamic simulation, we aim to
demonstrate how the internal structure of the depression--precariousness
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted exploratory analyses to identify
consistent and meaningful data-driven factor structures. These analyses
led to the identification of five distinct precariousness
factors---three domain-based and two reflecting recent stressors---each
composed of multiple observed variables, as outlined below. Additional
details on the HELIUS study and the factor extraction process are
provided in the \hyperref[sec-appendix]{Appendix}.

@@

After preprocessing, the HELIUS dataset includes 21,628 individuals. In
addition to the five identified precariousness factors, we computed an
overall precariousness score by by summing these five factor scores,
thereby capturing a composite measure of accumulated precariousness
across domains. Depression is represented using the PHQ-9 (Kroenke et
al., 2001), both as a total sum score and as individual symptom scores.
In the subsequent causal discovery analysis, we explore the relationship
between depression and precariousness using both aggregated scores and
their disaggregated components. See Figure~\ref{fig-dist} for the
overall distributions of the variables used in the analysis.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay. \emph{This shows the distributions of the variables used in the
causal analysis, plotted as histograms with overlaid density estimates.
The x-axis represents the value of each variable after preprocessing.
The y-axis shows the corresponding estimated probability density.}
\textbf{\emph{P.emp}} = \emph{employment precariousness};
\textbf{\emph{P.hou}} = \emph{housing precariousness};
\textbf{\emph{P.soc}} = \emph{social precariousness};
\textbf{\emph{S.fin}} = \emph{recent financial stressors};
\textbf{\emph{S.rel}} = \emph{recent relational stressors};
\textbf{\emph{precariousness}} = \emph{overall precariousness};
\textbf{\emph{PHQsum}} = \emph{PHQ-9 sum score}; \textbf{\emph{anh}} =
\emph{anhedonia}; \textbf{\emph{app}} = \emph{appetite};
\textbf{\emph{con}} = \emph{concentration}; \textbf{\emph{dep}} =
\emph{depressed mood}; \textbf{\emph{ene}} = \emph{energy};
\textbf{\emph{glt}} = \emph{guilty}; \textbf{\emph{mot}} = \emph{motor};
\textbf{\emph{slp}} = \emph{sleep}; \textbf{\emph{sui}} =
\emph{suicidal}.}

@@

\subsection{Modeling Dynamics between precariousness and
Depression}\label{modeling-dynamics-between-precariousness-and-depression}

@@

To explore optimal parameter configurations, we used the NSGA-II
algorithm as implemented in the \texttt{nsga2R} package (Franz \&
Nakamura, 2015). This algorithm is inspired by natural selection and
evolves a population of candidate solutions over multiple generations to
balance competing objectives. We began with a population of 500 randomly
generated parameter sets and allowed them to evolve over 300
generations. During this process, solutions were selected, combined, and
mutated using standard settings: a crossover probability of 0.9 (which
controls how often solutions are combined), a mutation probability of
0.25 (introducing random variation), and a tournament size of 2 (used to
pick the best candidates for reproduction). The result is a Pareto
front---a set of solutions that are considered optimal in the sense that
no single objective can be improved without worsening another. From this
diverse set of trade-off solutions (see Figure~\ref{fig-pareto}), we
selected the 30 parameter sets with the lowest overall error across
objectives. These top-performing configurations were then used in the
simulation phase to evaluate the potential effects of different
interventions.

@@

\subsection{Causal Structure Linking Precariousness and
Depression}\label{causal-structure-linking-precariousness-and-depression}

@@

The sum score graphs provide a high-level summary of how precariousness
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precariousness factors (\emph{P.hou},
\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
depression sum score (\emph{PHQsum}), as identified by the FCI
algorithm. As described in Section~\ref{sec-analysis}, dashed edges
indicate ties, and the edge endpoint at which the tie occurs is marked
by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the
relationships between precariousness factors and depression
(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
precariousness (\emph{P.soc}) are not identified as causes of
depression, whereas financial stress (\emph{S.fin}) appears to play a
causal role. While \emph{P.emp} and \emph{P.soc} are generally not
recognized as causes of other precariousness factors, \emph{S.fin}
emerges as a potential cause, as indicated by its circle edge endpoint.
Supporting this, the proportion matrix plot in panel (b) shows that
\emph{S.fin} has some probability of causally influencing either
\emph{P.emp} or \emph{P.soc.}

@@

Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
with depression and \emph{S.fin} through a latent confounder. Its
connection to \emph{P.soc} is less definitive---the proportion matrix in
Figure~\ref{fig-sum-2} indicates a considerable probability that this
relationship is either absent, ambiguous, or potentially bidirectional.
Meanwhile, housing precariousness (\emph{P.hou}) is not directly
causally related to depression but is linked to \emph{P.emp} and
\emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
causally influences \emph{P.emp} or if their relationship is mediated by
an unobserved confounder.

@@

\caption{\label{fig-sum}Resulting graph of precariousness factors and
depression sum score using FCI and proportion of edge endpoint types.}

@@

Despite these differences, some consistent patterns emerge across both
CI tests, particularly in the case of \emph{S.fin}, which appears to be
connected to nearly all other variables---though many of these
connections are marked by ties, reflecting uncertainty in
directionality. One exception is the stronger evidence suggesting that
\emph{S.fin} may cause changes in \emph{app}, while its relationships
with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
with depressive symptoms, with most edges pointing toward \emph{P.soc}
rather than outward from it. This suggests that depressive symptoms,
particularly \emph{dep} and \emph{glt}, may contribute to worsening
social precariousness rather than the other way around. Additionally,
\emph{anh} also shows some probability of causally influencing
\emph{P.soc}, reinforcing the idea that social precariousness is more
often a consequence rather than a driver of depressive symptoms.

@@

Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
to be the most connected to precariousness factors, while \emph{P.soc}
has the most connections with symptoms, predominantly as a recipient
rather than a driver of influence. Within the symptom network,
\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
symptoms, as they exhibit more outgoing arrows compared to other
symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
\emph{glt}, despite having high connectivity, predominantly receive
incoming arrows, indicating they are more likely effects rather than
causes. Considering both symptom-precariousness connections and
symptom-level dynamics, \emph{slp} emerges as a central symptom, given
its strong ties to precariousness factors and its influential role
within the symptom network. This suggests that sleep issues could be an
initiating symptom, particularly sensitive to relational stress and
employment precariousness. On the other hand, \emph{P.soc} serves as a
key bridge between the depression and precariousness subsystems, as
depressive symptoms appear to feed back into social precariousness,
reinforcing a self-sustaining dynamic between depression and precarious
conditions.

@@

\caption{\label{fig-sym}Resulting graph of precariousness factors and
individual depression symptoms using FCI and proportion of edge endpoint
types.}

@@

\subsubsection{Precariousness as sum
score}\label{precariousness-as-sum-score}

@@

Overall, the aggregated precariousness graph supports the key findings
from both the disaggregated analysis and the depression sum score
analysis, providing greater certainty in several important patterns. The
results highlight the significant role of latent confounders in
symptom-to-symptom interactions. While certain stressor-related
precariousness factors, particularly \emph{S.fin} and \emph{S.rel},
appear to causally influence depressive symptoms, other precariousness
factors are more likely driven by depression rather than acting as
primary causes, suggesting that precariousness may function more as a
consequence of depression than its source. Some symptoms, particularly
\emph{glt}, \emph{slp}, and \emph{dep}, appear to influence precarious
conditions while also occupying central roles within the symptom
network. Among them, \emph{slp} and \emph{glt} emerge as influential
symptoms within the depression network, whereas \emph{con} and
\emph{dep}, despite their high connectivity, behave more as outcomes
rather than primary drivers. Together, these findings suggest that
depression---especially symptoms like sleep disturbances and guilt---may
contribute to sustaining precarious conditions, possibly forming a
reciprocal cycle between depression and precariousness.

@@

\caption{\label{fig-presum}Resulting graph of precariousness sum score
and individual depression symptoms using FCI and proportion of edge
endpoint types.}

@@

\subsection{Simulating the Dynamics Between Precariousness and
Depression Under
Intervention}\label{simulating-the-dynamics-between-precariousness-and-depression-under-intervention}

To evaluate how internal system structure influences the effect of
external support, we simulated outcomes under a range of intervention
strengths across the top 30 parameter sets from the calibrated model
ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
levels of depression (\(D\)) and precariousness (\(P\)) after
simulation, with each line representing one parameter set. Lines are
color-coded by combined feedback and input strength, calculated as the
product \(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).

@@

Tailoring intervention strategies to the system's underlying structure
may enhance their effectiveness and sustainability, particularly in
addressing the compounding effects of socioeconomic precariousness on
mental health.

@@

\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
precariousness (\(P\)) across varying intervention strengths, simulated
over the top 30 parameter sets. Each line represents one parameter set
and is colored by its combined feedback and input strength.}
\textbf{Bottom}: Smoothed Density Heatmap of Final \(D\) and \(P\) for
Two Parameter Sets. \emph{One exhibits unistability (top two panels) and
the other exhibits bistability (bottom two panels). Bistable dynamics
reveal a sharp transition between distinct outcome regimes, while
unistable dynamics show gradual, continuous change across the
intervention gradient.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact---and how their internal structure modulates responsiveness to
intervention. Using cross-sectional data from the HELIUS study, we
applied cycle-capable discovery algorithms to estimate plausible
directional relationships among multiple domains of precariousness
factors, and depressive symptoms. By analyzing both aggregate constructs
and individual variables, we aimed to capture patterns that may be
masked in either extreme of abstraction (i.e., purely aggregate or
disaggregated analyses).

The causal analysis consistently identified financial stress as a key
plausible driver of depression, appearing across both aggregated and
disaggregated models. At the symptom level, we found that sleep
disturbance, depressed mood, guilt, and anhedonia were especially
sensitive to external stressors. Among these, the symptom depressed mood
emerged as a central hub: not only was it heavily influenced by other
symptoms, it also exerted directional influence on social
precariousness, suggesting a feedback loop in which worsening mood
contributes to social disconnection, which in turn may exacerbate
depressive symptoms. The CCI algorithm even suggested a possible cyclic
relationship between depressed mood and overall precariousness, hinting
at a structurally self-reinforcing mechanism. Symptom sleep disturbance
also played a key initiating role. It showed directed connections to
both social and employment precariousness and exhibited many outgoing
edges within the symptom network. This suggests the symptom sleep
disturbance may function as an early warning signal or entry point into
the depression--precariousness cycle, making it a promising target for
preventive intervention. In contrast, Social precariousness more often
appeared as an effect rather than a cause---receiving arrows from
symptoms like depressed mood and guilt, but rarely sending them. This
asymmetry implies that social precariousness may often reflect
downstream consequences of worsening mental health rather than being its
root cause. These patterns offer actionable insights for intervention.
Addressing symptoms like sleep disturbance early may prevent escalation
into broader depressive states or worsening precariousness. Intervening
on depressed mood, as a central node with potential feedback effects,
could also disrupt the self-reinforcing dynamics between mental health
and social isolation.

To examine how such structures shape responsiveness to support, we
developed a simplified two-variable nonlinear dynamical model capturing
the feedback between depression and precariousness, modulated by
financial stress. Calibrated against empirical summary statistics from
the HELIUS dataset, the model allowed us to simulate how these dynamics
evolve under varying levels of stress reduction. The simulations
revealed that intervention effectiveness depends not just on the
magnitude of external stressors, but on the internal configuration of
the system itself. In bistable regimes---characterized by strong
feedback and low noise---the system could remain stuck in a high-risk
state unless an intervention crosses a critical threshold, tipping it
into a healthier equilibrium. By contrast, unistable systems---driven by
weaker coupling or greater noise---responded more gradually and
proportionally to increasing support. These dynamics carry practical
policy implications. In bistable systems, even moderate interventions
may have little effect unless they are forceful enough to tip the system
toward a lower-risk state. This underscores the importance of early or
intensive support when targeting self-reinforcing cycles of depression
and precariousness. Unistable systems, by contrast, are more responsive
to incremental support, suggesting that sustained, moderate
interventions may be sufficient in such contexts.

@@

Finally, the structure of the dynamical model introduces its own
limitations. While the model was designed to reflect key features of the
discovered causal structure, it represents a highly simplified system:
only two variables---depression and social precariousness---modulated by
external financial stress. This abstraction allowed for tractable
analysis and visual intuition, particularly around nonlinear behaviors
such as bistability, but omits other potentially influential constructs
highlighted in the causal graphs, including relational stress,
employment conditions and housing conditions. Moreover, the model's
nonlinearity---introduced via the hyperbolic tangent function---was
chosen for convenience rather than empirical justification. While this
function supports saturation and the emergence of tipping points, the
presence of bistability is a product of simulation, not empirical
verification. It remains possible that simpler, linear dynamics could
account for the observed data just as well. Future work should further
investigate whether bistability is a necessary feature of
depression--precariousness dynamics, and explore alternative functional
forms---both linear and nonlinear---using richer data and stronger
empirical constraints.

Our results reveal that the effectiveness of interventions targeting
depression and precariousness is shaped not only by the strength of
external stressors but also by the internal structure of the system. The
interplay between feedback strength and noise determines whether a
system responds gradually or only after crossing a tipping
point---highlighting the need for intervention strategies attuned to
these underlying dynamics.
